---
title: "Forecasting and Volatility Modeling of Crude Oil Prices"
subtitle: "ADSP 31006 - Time Series Analysis and Forecasting | Final Project"
author: "Daniel Mendoza, Jared Maksoud, Nicholas Dhaliwal"
date: "March 9th, 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

**Course:** Time Series Analysis and Forecasting
**Instructor:** Prof. Arnab Bose, Ph.D.

---

# Project Overview

This project analyzes daily WTI (West Texas Intermediate, Cushing Oklahoma) and Brent
(North Sea) crude oil spot prices from the U.S. Energy Information Administration (EIA).
The dataset covers 1987 through early 2026, giving us roughly 9,700 trading-day
observations per series.

We have three objectives. First, forecast WTI price levels and log returns over a
63-trading-day horizon. Second, model time-varying conditional variance using ARCH/GARCH
methods. Third, measure the price impact of three geopolitical shocks using intervention
analysis.

**Contributions:**

- Daniel Mendoza: Section 3 (EDA)
- Jared Maksoud: Section 5 (Intervention Analysis)
- Nicholas Dhaliwal: Sections 4, 6, 7 (Feature Engineering, Forecasting Models, Comparison)

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width  = 9,
  fig.height = 5,
  fig.retina = 2
)
```

# Libraries

```{r libraries}
# Data wrangling and visualization
library(tidyverse)
library(lubridate)
library(readxl)
library(gridExtra)
library(knitr)
library(zoo)

# Univariate time series
library(forecast)   # auto.arima, ets, Arima, accuracy, checkresiduals
library(tseries)    # adf.test, kpss.test
library(urca)       # ca.po (Phillips-Ouliaris cointegration test)

# ARCH/GARCH
library(rugarch)    # ugarchspec, ugarchfit, ugarchforecast
library(FinTS)      # ArchTest (ARCH-LM test)

# Multivariate time series
library(vars)       # VAR, VARselect, causality, irf, predict

# rugarch and vars both pull in MASS as a dependency, which masks
# dplyr::select, dplyr::filter, and dplyr::lag. Re-assert dplyr versions
# here so downstream pipelines work correctly.
select <- dplyr::select
filter <- dplyr::filter
lag    <- dplyr::lag
```

---

# Data Loading and Cleaning

The EIA file is an Excel workbook with a metadata header above the actual data. The date
column mixes two formats: numeric Excel serial integers and ISO character strings. The
routine below handles both and locates the header row dynamically rather than hard-coding
a row number, which makes it robust to minor file layout changes.

```{r load-data}
x <- read_excel("PET_PRI_SPT_S1_D.xls", sheet = "Data 1", col_names = FALSE)

# Locate the header row dynamically
i   <- which(x[[1]] == "Date")[1]
dat <- x[(i + 1):nrow(x), 1:3]
names(dat) <- c("Date", "USCrudePrice", "EuroCrudePrice")

# Handle mixed date formats
dnum     <- suppressWarnings(as.numeric(dat$Date))
dat$Date <- ifelse(
  !is.na(dnum) & dnum > 20000,
  as.character(as.Date(dnum, origin = "1899-12-30")),
  as.character(dat$Date)
)
dat$Date <- as.Date(dat$Date, format = "%Y-%m-%d")
bad      <- is.na(dat$Date)
dat$Date[bad] <- as.Date(as.character(dat$Date[bad]), format = "%b %d, %Y")

# Coerce prices, drop incomplete rows, sort chronologically
oil <- dat |>
  mutate(
    USCrudePrice   = suppressWarnings(as.numeric(USCrudePrice)),
    EuroCrudePrice = suppressWarnings(as.numeric(EuroCrudePrice))
  ) |>
  filter(!is.na(Date), !is.na(USCrudePrice), !is.na(EuroCrudePrice)) |>
  arrange(Date)

cat("Date range  :", as.character(range(oil$Date)), "\n")
cat("Observations:", nrow(oil), "\n")
```

---

# Exploratory Data Analysis

*Section authored by Daniel Mendoza.*

We examine distributional properties, temporal dynamics, stationarity, autocorrelation
structure, and the long-run relationship between WTI and Brent.

## Price Series

```{r eda-price-plot}
ggplot(oil, aes(x = Date)) +
  geom_line(aes(y = USCrudePrice,   color = "WTI"),   linewidth = 0.5) +
  geom_line(aes(y = EuroCrudePrice, color = "Brent"), linewidth = 0.5) +
  scale_color_manual(values = c("WTI" = "#2c3e50", "Brent" = "#16a085")) +
  labs(title = "Crude Oil Daily Spot Prices: WTI and Brent",
       x = "Date", y = "Price (USD/bbl)", color = "Series") +
  theme_minimal(base_size = 12)
```

Both series trend together and show large dislocations around a few key events: the 2008
financial crisis, the 2014-16 OPEC supply glut, the 2020 COVID crash, and the 2022 Ukraine
invasion. These breaks motivate the intervention analysis in Section 5.

## Price Distributions

```{r eda-histograms, fig.width=10, fig.height=4}
p1 <- ggplot(oil, aes(x = USCrudePrice)) +
  geom_histogram(bins = 50, fill = "#2c3e50", color = "white", alpha = 0.85) +
  labs(title = "WTI Price Distribution", x = "Price (USD/bbl)", y = "Count") +
  theme_minimal()

p2 <- ggplot(oil, aes(x = EuroCrudePrice)) +
  geom_histogram(bins = 50, fill = "#16a085", color = "white", alpha = 0.85) +
  labs(title = "Brent Price Distribution", x = "Price (USD/bbl)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

Both distributions are right-skewed with long upper tails. Variance also increases with
the price level. Both of these properties motivate a log transformation before modeling.

## Log Transformation

```{r eda-log-transform}
oil <- oil |>
  mutate(
    logWTI   = log(USCrudePrice),
    logBrent = log(EuroCrudePrice)
  )

p3 <- ggplot(oil, aes(Date, logWTI)) +
  geom_line(color = "#2c3e50", linewidth = 0.5) +
  labs(title = "Log WTI Price", x = "Date", y = "log(Price)") +
  theme_minimal()

p4 <- ggplot(oil, aes(Date, logBrent)) +
  geom_line(color = "#16a085", linewidth = 0.5) +
  labs(title = "Log Brent Price", x = "Date", y = "log(Price)") +
  theme_minimal()

grid.arrange(p3, p4, ncol = 2)
```

The log transformation stabilizes variance across the sample and compresses the upper
tail. First differences of log prices give log returns, which are interpretable as
approximate percentage changes.

## Log Returns

```{r eda-log-returns}
oil <- oil |>
  arrange(Date) |>
  mutate(
    WTI_Return   = c(NA, diff(logWTI)),
    Brent_Return = c(NA, diff(logBrent))
  )

p5 <- ggplot(oil |> filter(!is.na(WTI_Return)), aes(Date, WTI_Return)) +
  geom_line(color = "#2c3e50", linewidth = 0.35) +
  labs(title = "WTI Log Returns", x = "Date", y = "Log Return") +
  theme_minimal()

p6 <- ggplot(oil |> filter(!is.na(Brent_Return)), aes(Date, Brent_Return)) +
  geom_line(color = "#16a085", linewidth = 0.35) +
  labs(title = "Brent Log Returns", x = "Date", y = "Log Return") +
  theme_minimal()

grid.arrange(p5, p6, ncol = 2)
```

Returns fluctuate around a stable mean near zero, consistent with a stationary series.
But the variance is clearly not constant over time.

## Volatility Clustering

```{r eda-vol-clustering, fig.width=10, fig.height=4}
p7 <- ggplot(oil |> filter(!is.na(WTI_Return)), aes(Date, abs(WTI_Return))) +
  geom_line(color = "purple", linewidth = 0.35) +
  labs(title = "WTI Absolute Returns", x = "Date", y = "|Log Return|") +
  theme_minimal()

p8 <- ggplot(oil |> filter(!is.na(Brent_Return)), aes(Date, abs(Brent_Return))) +
  geom_line(color = "purple", linewidth = 0.35) +
  labs(title = "Brent Absolute Returns", x = "Date", y = "|Log Return|") +
  theme_minimal()

grid.arrange(p7, p8, ncol = 2)
```

Large absolute returns cluster together and calm periods cluster together. This is the
signature of ARCH effects that we formally test before fitting GARCH.

## ACF and PACF of Returns

```{r eda-acf-pacf, fig.width=11, fig.height=6}
oil_clean <- oil |> filter(!is.na(WTI_Return), !is.na(Brent_Return))

par(mfrow = c(2, 2))
acf(oil_clean$WTI_Return,    lag.max = 40, main = "ACF - WTI Returns")
pacf(oil_clean$WTI_Return,   lag.max = 40, main = "PACF - WTI Returns")
acf(oil_clean$Brent_Return,  lag.max = 40, main = "ACF - Brent Returns")
pacf(oil_clean$Brent_Return, lag.max = 40, main = "PACF - Brent Returns")
par(mfrow = c(1, 1))
```

Most autocorrelations fall inside the 95% bands, consistent with near-white-noise behavior
in the conditional mean. A few marginally significant early lags suggest a low-order ARMA
may improve fit, which we let AICc resolve.

## Stationarity Tests

We use ADF and KPSS together because their null hypotheses are complementary. ADF tests
H0: unit root; KPSS tests H0: level stationarity. Agreement between the two gives
stronger evidence than either test alone.

```{r eda-stationarity}
adf_wti_lvl   <- adf.test(na.omit(oil$USCrudePrice))
adf_brent_lvl <- adf.test(na.omit(oil$EuroCrudePrice))
adf_wti_ret   <- adf.test(na.omit(oil$WTI_Return))
adf_brent_ret <- adf.test(na.omit(oil$Brent_Return))

kpss_wti_lvl   <- kpss.test(na.omit(oil$USCrudePrice))
kpss_brent_lvl <- kpss.test(na.omit(oil$EuroCrudePrice))
kpss_wti_ret   <- kpss.test(na.omit(oil$WTI_Return))
kpss_brent_ret <- kpss.test(na.omit(oil$Brent_Return))

stationarity_tbl <- data.frame(
  Series    = c("WTI Levels", "Brent Levels", "WTI Returns", "Brent Returns"),
  ADF_p     = round(c(adf_wti_lvl$p.value, adf_brent_lvl$p.value,
                      adf_wti_ret$p.value, adf_brent_ret$p.value), 4),
  ADF_conc  = c("Non-stationary", "Non-stationary", "Stationary", "Stationary"),
  KPSS_p    = round(c(kpss_wti_lvl$p.value, kpss_brent_lvl$p.value,
                       kpss_wti_ret$p.value, kpss_brent_ret$p.value), 4),
  KPSS_conc = c("Non-stationary", "Non-stationary", "Stationary", "Stationary")
)

kable(stationarity_tbl,
      col.names = c("Series", "ADF p-val", "ADF Result", "KPSS p-val", "KPSS Result"),
      caption   = "Table 1: Stationarity Test Results")
```

Both tests agree. Price levels are I(1) and log returns are I(0). This sets d = 0 for
all return-based models and d = 1 for any level-based model.

## Correlation and Cointegration

```{r eda-correlation}
r_price <- cor(oil$USCrudePrice, oil$EuroCrudePrice, use = "complete.obs")
r_ret   <- cor(oil$WTI_Return,   oil$Brent_Return,   use = "complete.obs")
cat("Correlation (price levels):", round(r_price, 4), "\n")
cat("Correlation (log returns) :", round(r_ret,   4), "\n")

ggplot(oil, aes(USCrudePrice, EuroCrudePrice)) +
  geom_point(alpha = 0.12, color = "#2c3e50", size = 0.7) +
  geom_smooth(method = "lm", color = "#16a085", se = FALSE, linewidth = 1) +
  labs(title = paste0("WTI vs. Brent (r = ", round(r_price, 3), ")"),
       x = "WTI Price (USD/bbl)", y = "Brent Price (USD/bbl)") +
  theme_minimal()
```

```{r eda-cointegration}
# Phillips-Ouliaris test. H0: no cointegration.
coint_test <- ca.po(
  cbind(oil$USCrudePrice, oil$EuroCrudePrice),
  demean = "constant"
)
summary(coint_test)
```

```{r eda-spread}
oil <- oil |> mutate(spread = USCrudePrice - EuroCrudePrice)

ggplot(oil, aes(Date, spread)) +
  geom_line(color = "darkgreen", linewidth = 0.5) +
  geom_hline(yintercept = mean(oil$spread, na.rm = TRUE),
             linetype = "dashed", color = "gray50") +
  labs(title = "WTI-Brent Spot Price Spread",
       x = "Date", y = "Spread (USD/bbl)") +
  theme_minimal()

cat("Mean spread:", round(mean(oil$spread, na.rm = TRUE), 2), "USD/bbl\n")
cat("SD spread  :", round(sd(oil$spread,   na.rm = TRUE), 2), "USD/bbl\n")
```

The Phillips-Ouliaris test rejects no cointegration. WTI and Brent share a long-run
equilibrium even though each is individually I(1). The spread is therefore I(0) and
mean-reverting. This supports two modeling choices: the spread is a usable stationary
feature, and a VAR on the joint return system is appropriate.

**EDA Summary:**

| Finding | Modeling Implication |
|---------|---------------------|
| Prices I(1), returns I(0) | d = 0 for return models, d = 1 for level models |
| Volatility clustering in abs. returns | ARCH effects warrant a GARCH model |
| Returns near white noise in the mean | Low-order ARMA likely sufficient |
| Correlation r ~0.99, cointegrated | VAR on joint system appropriate |

---

# Feature Engineering

*Group contribution, motivated by EDA findings.*

Each feature below was built for a specific reason tied to the EDA. The comments explain
the connection to course material.

```{r feature-engineering}
oil <- oil |>
  arrange(Date) |>
  mutate(

    # Log returns: first difference of log prices.
    # Converts I(1) prices to I(0) returns (differencing from class).
    WTI_Return   = c(NA, diff(logWTI)),
    Brent_Return = c(NA, diff(logBrent)),

    # WTI-Brent spread.
    # Cointegration in Section 3 implies this linear combination is I(0),
    # so the spread is stationary and mean-reverting. It could serve as
    # an xreg in a dynamic regression if we extended the ARIMA model.
    spread = USCrudePrice - EuroCrudePrice,

    # 10-day rolling realized volatility.
    # A simple non-parametric variance proxy. We use it to determine
    # whether each geopolitical shock should be modeled as a pulse or
    # step (Section 5). A 10-day window is roughly two trading weeks.
    rolling_vol_10 = rollapply(
      WTI_Return, width = 10, FUN = sd, fill = NA, align = "right"
    ),

    # Geopolitical shock dummies, following the intervention analysis
    # framework from class (pulse vs. step).
    # Pulse P_t: 1 during shock window, 0 otherwise. COVID gets a pulse
    #   because rolling vol resolved to baseline by ~July 2020.
    # Step S_t: 1 from event date onward. Ukraine and Middle East get
    #   steps because rolling vol stayed elevated with no reversion.
    covid_pulse     = as.integer(
      Date >= as.Date("2020-03-13") & Date <= as.Date("2020-07-01")
    ),
    ukraine_step    = as.integer(Date >= as.Date("2022-02-24")),
    middleeast_step = as.integer(Date >= as.Date("2023-10-06")),

    # Squared returns as a realized variance proxy.
    # Used in the McLeod-Li test (Ljung-Box on r_t^2 to detect ARCH
    # effects) and as the out-of-sample evaluation target for GARCH.
    WTI_sq_return = WTI_Return^2
  )

feat_tbl <- data.frame(
  Feature = c("WTI_Return", "Brent_Return", "spread",
              "rolling_vol_10", "covid_pulse",
              "ukraine_step", "middleeast_step", "WTI_sq_return"),
  Type    = c("Continuous", "Continuous", "Continuous",
              "Continuous", "Binary", "Binary", "Binary", "Continuous"),
  Order   = c("I(0)", "I(0)", "I(0)", "I(0)",
              "I(0)", "I(0)", "I(0)", "I(0)"),
  Purpose = c(
    "Stationarity via differencing (ARIMA, GARCH, VAR)",
    "Stationarity via differencing (VAR)",
    "I(0) spread from cointegrated pair",
    "Rolling vol for pulse/step classification",
    "Pulse dummy: COVID temporary shock",
    "Step dummy: Ukraine permanent shift",
    "Step dummy: Middle East permanent shift",
    "Realized variance for McLeod-Li and GARCH eval"
  )
)

kable(feat_tbl,
      col.names = c("Feature", "Type", "Order", "Purpose"),
      caption   = "Table 2: Engineered Features")
```

---

# Train / Test Split

The most recent 63 trading days (~3 calendar months) are held out as the test set. All
four forecasting models are trained on data prior to this window and evaluated on it.
This avoids any data leakage and gives a fair comparison across models.

```{r train-test}
h_test <- 63

oil_ret   <- oil |> filter(!is.na(WTI_Return), !is.na(Brent_Return))
n_ret     <- nrow(oil_ret)
train_ret <- oil_ret[1:(n_ret - h_test), ]
test_ret  <- oil_ret[(n_ret - h_test + 1):n_ret, ]

n_lvl     <- nrow(oil)
train_lvl <- oil[1:(n_lvl - h_test), ]
test_lvl  <- oil[(n_lvl - h_test + 1):n_lvl, ]

cat("Training obs:", nrow(train_ret), "\n")
cat("Test obs    :", nrow(test_ret),  "\n")
cat("Train range :", as.character(range(train_ret$Date)), "\n")
cat("Test range  :", as.character(range(test_ret$Date)),  "\n")
```

---

# Intervention Analysis

*Section authored by Jared Maksoud.*

Before building forecasting models we identify structural breaks caused by three
geopolitical shocks: the COVID-19 crash (March 2020), the Russia-Ukraine invasion
(February 2022), and the Middle East escalation (October 2023).

From the class framework, an intervention can take two forms. A pulse (P_t) is a temporary
disruption: the dummy equals 1 during the shock window and 0 elsewhere. A step (S_t) is
a permanent level shift: the dummy equals 1 from the event date onward. We determine
which form applies to each event empirically using rolling volatility plots, not by
assumption.

## Event Timeline

```{r int-timeline}
plot(oil$Date, oil$USCrudePrice,
     type = "l", col = "black", lwd = 0.7,
     xlab = "Date", ylab = "WTI Price (USD/bbl)",
     main = "WTI Price with Geopolitical Event Dates")
abline(v = as.Date("2020-03-13"), col = "blue",   lty = 2, lwd = 1.5)
abline(v = as.Date("2022-02-24"), col = "red",    lty = 2, lwd = 1.5)
abline(v = as.Date("2023-10-06"), col = "purple", lty = 2, lwd = 1.5)
legend("topleft",
       legend = c("COVID (Mar 2020)",
                  "Ukraine (Feb 2022)",
                  "Middle East (Oct 2023)"),
       col = c("blue", "red", "purple"), lty = 2, bty = "n", cex = 0.85)
```

## Rolling Volatility Around Each Event

```{r int-rolling-vol, fig.height=9}
oil2 <- oil |> filter(!is.na(USCrudePrice), USCrudePrice > 0)

covid_win <- oil2 |>
  filter(Date >= as.Date("2020-01-01"), Date <= as.Date("2020-12-31")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

ukraine_win <- oil2 |>
  filter(Date >= as.Date("2021-12-01"), Date <= as.Date("2022-12-01")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

me_win <- oil2 |>
  filter(Date >= as.Date("2023-06-01"), Date <= as.Date("2024-06-01")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

par(mfrow = c(3, 1), mar = c(3, 4, 3, 1))

plot(covid_win$Date, covid_win$vol, type = "l", col = "black",
     xlab = "", ylab = "10-day rolling sd",
     main = "COVID (2020): volatility resolved ~Jul 2020, treated as pulse")
abline(v = as.Date("2020-03-11"), col = "blue", lty = 2, lwd = 1.5)
abline(v = as.Date("2020-07-01"), col = "blue", lty = 3, lwd = 1.5)

plot(ukraine_win$Date, ukraine_win$vol, type = "l", col = "black",
     xlab = "", ylab = "10-day rolling sd",
     main = "Ukraine (2022): volatility stayed elevated, treated as step")
abline(v = as.Date("2022-02-24"), col = "red", lty = 2, lwd = 1.5)

plot(me_win$Date, me_win$vol, type = "l", col = "black",
     xlab = "Date", ylab = "10-day rolling sd",
     main = "Middle East (2023-24): volatility did not revert, treated as step")
abline(v = as.Date("2023-10-06"), col = "purple", lty = 2, lwd = 1.5)

par(mfrow = c(1, 1))
```

COVID volatility resolved by approximately July 2020, consistent with a pulse. Ukraine
and Middle East volatility did not revert within the observation window, consistent with
permanent step shifts.

## Intervention ARIMA

We fit ARIMA with the three shock dummies as external regressors. This is the
`auto.arima(y, xreg = X)` dynamic regression approach from class, extended with the
intervention dummy encoding. The model is estimated on the full series because the goal
here is to estimate the shock coefficients, not to produce out-of-sample forecasts.

```{r int-arima-fit}
y_int <- ts(oil2$USCrudePrice, frequency = 252)

covid_d   <- as.integer(oil2$Date >= as.Date("2020-03-13") &
                        oil2$Date <= as.Date("2020-07-01"))
ukraine_d <- as.integer(oil2$Date >= as.Date("2022-02-24"))
me_d      <- as.integer(oil2$Date >= as.Date("2023-10-06"))

X_int <- cbind(covid        = covid_d,
               ukraine_step = ukraine_d,
               middleeast   = me_d)

cat("Observations per dummy:\n")
print(colSums(X_int))

fit_int <- auto.arima(
  y_int,
  xreg          = X_int,
  seasonal      = FALSE,
  stepwise      = FALSE,
  approximation = FALSE
)

summary(fit_int)
checkresiduals(fit_int)
```

```{r int-coefs}
beta_int <- coef(fit_int)[c("covid", "ukraine_step", "middleeast")]

cat("\nIntervention Coefficient Estimates:\n")
cat("  COVID pulse effect      :", round(beta_int["covid"],        2),
    "USD/bbl (Mar 13 - Jul 1, 2020)\n")
cat("  Ukraine step effect     :", round(beta_int["ukraine_step"], 2),
    "USD/bbl (permanent from Feb 24, 2022)\n")
cat("  Middle East step effect :", round(beta_int["middleeast"],   2),
    "USD/bbl (permanent from Oct 6, 2023)\n")

int_covid_coef <- round(beta_int["covid"],        2)
int_ukr_coef   <- round(beta_int["ukraine_step"], 2)
int_me_coef    <- round(beta_int["middleeast"],   2)
```

## Counterfactual: Actual vs. Fitted vs. No Interventions

```{r int-counterfactual}
yhat_int   <- as.numeric(fitted(fit_int))
shock_part <- as.numeric(X_int %*% beta_int)
yhat_cf    <- yhat_int - shock_part

plot(oil2$Date, as.numeric(y_int),
     type = "l", col = "black", lwd = 0.7,
     xlab = "Date", ylab = "WTI (USD/bbl)",
     main = "WTI: Actual vs. Fitted vs. No-Intervention Counterfactual")
lines(oil2$Date, yhat_int, col = "blue",      lwd = 1.8)
lines(oil2$Date, yhat_cf,  col = "darkgreen", lwd = 1.8, lty = 2)
abline(v = as.Date("2020-03-13"), col = "gray50", lty = 3)
abline(v = as.Date("2022-02-24"), col = "gray50", lty = 3)
abline(v = as.Date("2023-10-06"), col = "gray50", lty = 3)
legend("topleft",
       legend = c("Actual", "Fitted (with interventions)",
                  "Counterfactual (no interventions)"),
       col = c("black", "blue", "darkgreen"),
       lty = c(1, 1, 2), bty = "n", cex = 0.85)
```

```{r int-contribution}
plot(oil2$Date, shock_part,
     type = "l", col = "darkred", lwd = 1,
     xlab = "Date", ylab = "Intervention contribution (USD/bbl)",
     main = "Total Estimated Intervention Effect Over Time")
abline(h = 0, lty = 2, col = "gray60")
abline(v = as.Date("2020-03-13"), col = "blue",   lty = 3)
abline(v = as.Date("2022-02-24"), col = "red",    lty = 3)
abline(v = as.Date("2023-10-06"), col = "purple", lty = 3)
```

---

# Forecasting Models

*Section authored by Nicholas Dhaliwal.*

Four models are developed, each motivated by a specific EDA finding. Together with the
intervention ARIMA above, the project covers five distinct models.

| Model | EDA Finding | Class Coverage |
|-------|------------|----------------|
| ARIMA | Returns I(0), near-WN ACF/PACF | Week 3 |
| ETS | Prices trend, no fixed seasonality | Week 2 |
| ARMA-GARCH | Volatility clustering, ARCH confirmed | Week 7 |
| VAR | r ~0.99, cointegration confirmed | Week 7 |

## ARIMA on WTI Log Returns

Returns are I(0) so we fit ARIMA(p, 0, q). The Box-Jenkins procedure from class guides
identification: inspect ACF/PACF, run an exhaustive AICc search, then compare neighboring
models to confirm the selection is stable. Setting d = 0 prevents auto.arima from
differencing a series that is already stationary.

### ACF and PACF

```{r arima-acf}
par(mfrow = c(1, 2))
acf(train_ret$WTI_Return,  lag.max = 40, main = "ACF - WTI Returns (train)")
pacf(train_ret$WTI_Return, lag.max = 40, main = "PACF - WTI Returns (train)")
par(mfrow = c(1, 1))
```

### Model Selection

```{r arima-fit}
wti_ret_ts <- ts(train_ret$WTI_Return, frequency = 252)

arima_fit <- auto.arima(
  wti_ret_ts,
  d             = 0,        # returns are I(0), do not difference
  max.p         = 5,
  max.q         = 5,
  stepwise      = FALSE,    # exhaustive search
  approximation = FALSE,    # exact likelihood
  ic            = "aicc"    # AICc penalizes overfitting more than AIC
)

summary(arima_fit)

p_ar <- arima_fit$arma[1]
q_ma <- arima_fit$arma[2]
cat("\nauto.arima selected: ARIMA(", p_ar, ",0,", q_ma, ")\n", sep = "")
cat("AICc              :", round(arima_fit$aicc,   3), "\n")
cat("Log-likelihood    :", round(arima_fit$loglik, 3), "\n")
```

### Neighboring Model Comparison

```{r arima-neighbors}
cands <- list(
  selected = arima_fit,
  p_up     = Arima(wti_ret_ts, order = c(p_ar + 1, 0, q_ma)),
  q_up     = Arima(wti_ret_ts, order = c(p_ar,     0, q_ma + 1))
)
if (p_ar > 0) cands$p_down <- Arima(wti_ret_ts, order = c(p_ar - 1, 0, q_ma))
if (q_ma > 0) cands$q_down <- Arima(wti_ret_ts, order = c(p_ar,     0, q_ma - 1))

nbr_tbl <- data.frame(
  Model = names(cands),
  AICc  = round(sapply(cands, function(m) m$aicc), 3)
) |> arrange(AICc)

kable(nbr_tbl, caption = "Table 3: ARIMA Neighboring Model AICc Comparison")
```

### Diagnostics

```{r arima-diagnostics}
checkresiduals(arima_fit)
```

```{r arima-ljungbox}
lb_lags <- c(5, 10, 20)
lb_tbl  <- data.frame(Lag = lb_lags, Q_stat = NA_real_, p_value = NA_real_)
for (k in seq_along(lb_lags)) {
  lb <- Box.test(residuals(arima_fit), lag = lb_lags[k],
                 type = "Ljung-Box", fitdf = p_ar + q_ma)
  lb_tbl$Q_stat[k]  <- round(lb$statistic, 3)
  lb_tbl$p_value[k] <- round(lb$p.value,   4)
}
kable(lb_tbl, caption = "Table 4: Ljung-Box Test on ARIMA Residuals")
```

### Forecast and Accuracy

```{r arima-forecast}
arima_fcast <- forecast(arima_fit, h = h_test)

autoplot(arima_fcast) +
  labs(title = paste0("ARIMA(", p_ar, ",0,", q_ma, ") - WTI Return Forecast (",
                      h_test, " days)"),
       x = "Time (training days)", y = "Log Return") +
  theme_minimal()
```

```{r arima-accuracy}
arima_acc <- accuracy(arima_fcast, test_ret$WTI_Return)
kable(round(arima_acc, 6), caption = "Table 5: ARIMA Accuracy")

arima_rmse <- arima_acc["Test set", "RMSE"]
arima_mae  <- arima_acc["Test set", "MAE"]
arima_mase <- arima_acc["Test set", "MASE"]
```

The point forecast converges to zero quickly, which is expected for a near-random-walk
return series. The value of ARIMA here is the calibrated prediction intervals and the
pre-whitened residuals, which feed into the GARCH pre-test below.

---

## ETS on WTI Price Levels

ETS models price levels directly without a stationarity transformation. The `ets()`
function searches all error/trend/season combinations and picks the minimum-AICc
specification, the same information criterion used for ARIMA. Given that oil prices trend
persistently without obvious calendar seasonality, we expect a trend component and no
seasonal component.

```{r ets-fit}
wti_lvl_ts <- ts(train_lvl$USCrudePrice, frequency = 252)

ets_fit <- ets(wti_lvl_ts, restrict = FALSE)

summary(ets_fit)
cat("\nSelected:", ets_fit$method, "\n")
cat("AICc    :", round(ets_fit$aicc, 2), "\n")
```

### Diagnostics

```{r ets-diagnostics}
checkresiduals(ets_fit)
```

### Forecast and Accuracy

```{r ets-forecast}
ets_fcast <- forecast(ets_fit, h = h_test)

autoplot(ets_fcast) +
  labs(title = paste0("ETS(", ets_fit$method, ") - WTI Price Level Forecast (",
                      h_test, " days)"),
       x = "Time (training days)", y = "WTI Price (USD/bbl)") +
  theme_minimal()
```

```{r ets-accuracy}
ets_acc <- accuracy(ets_fcast, test_lvl$USCrudePrice)
kable(round(ets_acc, 4), caption = "Table 6: ETS Accuracy")

ets_rmse <- ets_acc["Test set", "RMSE"]
ets_mae  <- ets_acc["Test set", "MAE"]
ets_mase <- ets_acc["Test set", "MASE"]
```

---

## ARMA-GARCH Volatility Model

Two EDA findings motivate this model. First, absolute returns show clear volatility
clustering. Second, ACF of squared returns shows significant autocorrelation. Standard
ARIMA assumes homoscedastic errors and cannot address either. ARMA-GARCH decouples the
problem: ARMA handles the conditional mean and GARCH(1,1) models the conditional variance.

We use GARCH(1,1) because it is equivalent to an infinite-order ARCH model with only two
variance parameters (alpha1 and beta1). Student-t innovations accommodate the heavy tails
visible in the return histograms.

### Pre-Test: ARCH Effects

```{r garch-arch-tests}
arima_resid <- residuals(arima_fit)

# McLeod-Li: Ljung-Box on squared residuals
ml_lags <- c(5, 10, 20)
ml_tbl  <- data.frame(Lag = ml_lags, Q_squared = NA_real_, p_value = NA_real_)
for (k in seq_along(ml_lags)) {
  lb_sq <- Box.test(arima_resid^2, lag = ml_lags[k], type = "Ljung-Box")
  ml_tbl$Q_squared[k] <- round(lb_sq$statistic, 3)
  ml_tbl$p_value[k]   <- round(lb_sq$p.value,   6)
}
cat("McLeod-Li Test (Ljung-Box on squared residuals):\n")
kable(ml_tbl, caption = "Table 7: McLeod-Li Test Results")

# ARCH-LM test
arch_lm <- ArchTest(arima_resid, lags = 10)
cat("\nARCH-LM Test:\n")
print(arch_lm)
```

Both tests reject no ARCH effects. GARCH modeling is warranted.

### Model Fit

```{r garch-fit}
garch_spec <- ugarchspec(
  variance.model     = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model         = list(armaOrder = c(p_ar, q_ma), include.mean = TRUE),
  distribution.model = "std"   # Student-t for heavy tails
)

garch_fit <- ugarchfit(
  spec   = garch_spec,
  data   = train_ret$WTI_Return,
  solver = "hybrid"
)

show(garch_fit)
```

```{r garch-persistence}
alpha1      <- coef(garch_fit)["alpha1"]
beta1       <- coef(garch_fit)["beta1"]
shape_nu    <- coef(garch_fit)["shape"]
persistence <- alpha1 + beta1
half_life   <- log(0.5) / log(persistence)

cat("\nGARCH(1,1) Key Parameters:\n")
cat("  alpha1 (shock impact) :", round(alpha1,      5), "\n")
cat("  beta1  (persistence)  :", round(beta1,       5), "\n")
cat("  alpha1 + beta1        :", round(persistence, 5), "\n")
cat("  half-life of vol shock:", round(half_life,   1), "trading days\n")
cat("  Student-t dof (nu)    :", round(shape_nu,    2), "\n")

garch_persistence <- round(persistence, 4)
garch_halflife    <- round(half_life,   1)
garch_alpha1      <- round(alpha1, 5)
garch_beta1       <- round(beta1,  5)
```

Persistence (alpha1 + beta1) close to 1 means volatility shocks decay slowly. The
half-life is the number of trading days it takes for a shock to decay by half.

### Diagnostics

```{r garch-diagnostics, fig.width=11, fig.height=6}
std_resid <- residuals(garch_fit, standardize = TRUE)

par(mfrow = c(2, 2))
acf(std_resid,    lag.max = 30, main = "ACF - Standardized Residuals")
acf(std_resid^2,  lag.max = 30, main = "ACF - Squared Std. Residuals")
pacf(std_resid,   lag.max = 30, main = "PACF - Standardized Residuals")
pacf(std_resid^2, lag.max = 30, main = "PACF - Squared Std. Residuals")
par(mfrow = c(1, 1))
```

```{r garch-lb}
lb_mean_p <- Box.test(std_resid,   lag = 10, type = "Ljung-Box")$p.value
lb_var_p  <- Box.test(std_resid^2, lag = 10, type = "Ljung-Box")$p.value

cat("LB on std. residuals (mean adequacy) : p =", round(lb_mean_p, 4), "\n")
cat("LB on sq. std. residuals (var adequacy): p =", round(lb_var_p,  4), "\n")
```

### Conditional Volatility

```{r garch-cond-vol}
cond_vol_hist <- sigma(garch_fit)

plot(train_ret$Date, cond_vol_hist,
     type = "l", col = "darkred", lwd = 0.7,
     xlab = "Date", ylab = "Conditional volatility (sigma_t)",
     main = "GARCH(1,1) Estimated Conditional Volatility - Training Set")
abline(v = as.Date("2008-09-15"), col = "orange", lty = 2)
abline(v = as.Date("2020-03-13"), col = "blue",   lty = 2)
abline(v = as.Date("2022-02-24"), col = "red",    lty = 2)
abline(v = as.Date("2023-10-06"), col = "purple", lty = 2)
legend("topright",
       legend = c("sigma_t", "Lehman (2008)", "COVID (2020)",
                  "Ukraine (2022)", "ME (2023)"),
       col = c("darkred", "orange", "blue", "red", "purple"),
       lty = c(1, 2, 2, 2, 2), bty = "n", cex = 0.8)
```

### Volatility Forecast

```{r garch-forecast}
garch_fcast  <- ugarchforecast(garch_fit, n.ahead = h_test)
fcast_vol    <- as.numeric(sigma(garch_fcast))
realized_var <- test_ret$WTI_sq_return

fcast_dates_g <- seq(max(train_ret$Date) + 1, by = "day", length.out = h_test)

plot(fcast_dates_g, fcast_vol,
     type = "l", col = "darkred", lwd = 2,
     xlab = "Date", ylab = "Forecast sigma_t",
     main = paste0("GARCH(1,1) ", h_test, "-Day Volatility Forecast"))

vol_rmse <- sqrt(mean((fcast_vol^2 - realized_var)^2, na.rm = TRUE))
vol_mae  <-      mean( abs(fcast_vol^2 - realized_var),  na.rm = TRUE)

cat("GARCH forecast accuracy (forecast sigma^2 vs. realized r^2):\n")
cat("  RMSE:", round(vol_rmse, 8), "\n")
cat("  MAE :", round(vol_mae,  8), "\n")
```

---

## VAR on WTI and Brent Returns

WTI and Brent have correlation ~0.99 and are cointegrated. Modeling WTI alone ignores
information in Brent's lagged returns. VAR(p) models both series jointly and lets each
help forecast the other. We then test Granger causality to check whether that information
is statistically meaningful.

Note: because prices are I(1) and cointegrated, a VECM would be more efficient by imposing
the long-run equilibrium constraint. VAR on I(0) returns is valid but discards that
channel. We flag VECM as a natural extension.

### Stationarity Check

```{r var-stationarity}
cat("ADF on training returns:\n")
cat("  WTI  : p =", round(adf.test(train_ret$WTI_Return)$p.value,   4), "\n")
cat("  Brent: p =", round(adf.test(train_ret$Brent_Return)$p.value, 4), "\n")
```

### Lag Order Selection

```{r var-lag}
ret_mat <- cbind(WTI   = train_ret$WTI_Return,
                 Brent = train_ret$Brent_Return)

var_sel <- VARselect(ret_mat, lag.max = 15, type = "const")
cat("Lag order by criterion:\n")
print(var_sel$selection)

p_var <- var_sel$selection["AIC(n)"]
cat("\nUsing AIC-selected lag p =", p_var, "\n")
```

### Model Estimation

```{r var-fit}
var_fit <- VAR(ret_mat, p = p_var, type = "const")
summary(var_fit)
```

### Granger Causality

```{r var-granger}
gc_b2w <- causality(var_fit, cause = "Brent")
gc_w2b <- causality(var_fit, cause = "WTI")

cat("Granger Causality: Brent -> WTI\n")
print(gc_b2w$Granger)
cat("\nGranger Causality: WTI -> Brent\n")
print(gc_w2b$Granger)

gc_b2w_p <- round(gc_b2w$Granger$p.value, 4)
gc_w2b_p <- round(gc_w2b$Granger$p.value, 4)

cat("\nBrent -> WTI  : p =", gc_b2w_p,
    ifelse(gc_b2w_p < 0.05, "-> Granger-causes WTI", "-> no G-causality"), "\n")
cat("WTI -> Brent  : p =", gc_w2b_p,
    ifelse(gc_w2b_p < 0.05, "-> Granger-causes Brent", "-> no G-causality"), "\n")
```

### Impulse Response Functions

```{r var-irf, fig.width=11, fig.height=5}
irf_b2w <- irf(var_fit, impulse = "Brent", response = "WTI",
               n.ahead = 15, boot = TRUE, ci = 0.95)
irf_w2b <- irf(var_fit, impulse = "WTI",   response = "Brent",
               n.ahead = 15, boot = TRUE, ci = 0.95)

par(mfrow = c(1, 2))
plot(irf_b2w, main = "Brent shock -> WTI response (95% CI)")
plot(irf_w2b, main = "WTI shock -> Brent response (95% CI)")
par(mfrow = c(1, 1))
```

### Forecast and Accuracy

```{r var-forecast}
var_fcast  <- predict(var_fit, n.ahead = h_test)
wti_var_pt <- var_fcast$fcst$WTI[, "fcst"]
wti_var_lo <- var_fcast$fcst$WTI[, "lower"]
wti_var_hi <- var_fcast$fcst$WTI[, "upper"]

fcast_dates_v <- seq(max(train_ret$Date) + 1, by = "day", length.out = h_test)

plot(fcast_dates_v, wti_var_pt,
     type = "l", col = "steelblue", lwd = 2,
     ylim = range(c(wti_var_lo, wti_var_hi)),
     xlab = "Date", ylab = "WTI Log Return",
     main = paste0("VAR(", p_var, ") - WTI Return Forecast (", h_test, " days)"))
polygon(c(fcast_dates_v, rev(fcast_dates_v)),
        c(wti_var_hi, rev(wti_var_lo)),
        col = adjustcolor("steelblue", 0.15), border = NA)
abline(h = 0, lty = 2, col = "gray60")
```

```{r var-accuracy}
var_acc <- accuracy(wti_var_pt, test_ret$WTI_Return)
kable(round(var_acc, 6), caption = "Table 8: VAR Accuracy - WTI Returns (Test Set)")

var_rmse <- var_acc[1, "RMSE"]
var_mae  <- var_acc[1, "MAE"]
var_mase <- var_acc[1, "MASE"]
```

---

# Model Comparison

```{r comparison}
comp_tbl <- data.frame(
  Model  = c("ARIMA", "ETS", "ARMA-GARCH", "VAR", "Intervention ARIMA*"),
  Target = c("WTI log returns", "WTI price (USD/bbl)",
             "Conditional volatility", "WTI + Brent returns",
             "WTI price with shock dummies"),
  RMSE   = c(round(arima_rmse, 6), round(ets_rmse, 4),
             round(vol_rmse,   8), round(var_rmse, 6), NA),
  MAE    = c(round(arima_mae,  6), round(ets_mae,  4),
             round(vol_mae,    8), round(var_mae,  6), NA),
  MASE   = c(round(arima_mase, 4), round(ets_mase, 4),
             NA,                   round(var_mase, 4), NA)
)

kable(comp_tbl,
      col.names = c("Model", "Target", "RMSE", "MAE", "MASE"),
      caption   = "Table 9: Model Comparison",
      na        = "--")
```

*Intervention ARIMA is fitted on the full series for structural estimation. Out-of-sample
accuracy is not applicable.*

ARIMA and VAR both target WTI log returns and can be compared directly on RMSE and MAE.
ETS targets price levels so its error metrics are on a different scale and are not
directly comparable. GARCH targets conditional variance and is evaluated against realized
squared returns.

The recommended approach by use case: for **volatility or risk management**, ARMA-GARCH
is the primary model since it is the only one that explicitly forecasts variance. For
**return forecasting**, compare VAR vs. ARIMA directly. For **price level forecasting**,
use ETS. For **shock attribution**, the intervention ARIMA coefficients quantify how much
each geopolitical event shifted the WTI price level.

---

# Assumptions and Limitations

1. **Trading-day gaps.** Weekends and holidays are structural absences, not missing data.
   We use frequency = 252 throughout as the standard trading-day approximation.

2. **GARCH distribution.** Student-t handles heavy tails but assumes a symmetric
   distribution. A skewed-t could be tested as a robustness check.

3. **VAR vs. VECM.** VAR on I(0) returns is valid but ignores the long-run
   error-correction relationship confirmed by Phillips-Ouliaris. VECM would be more
   efficient at longer horizons and is a natural next step.

4. **Intervention dummy design.** Pulse and step classification is based on visual
   inspection of rolling volatility, not a formal change-point test. The COVID end date
   (July 1, 2020) involves some judgment.

5. **Unmodeled breaks.** The intervention dummies only capture three known events. Other
   structural breaks (e.g., the 2014-16 OPEC supply war) are not explicitly modeled and
   will inflate variance estimates.

---

# Future Work

```{r future-work}
fw_tbl <- data.frame(
  Extension   = c("VECM", "Regime-switching GARCH", "TimeGPT"),
  Rationale   = c(
    "Cointegration is confirmed; VECM imposes the long-run constraint and should improve multi-step accuracy over VAR",
    "Model distinct calm and crisis volatility regimes that the intervention analysis identified empirically",
    "Foundation model benchmark from class; compare zero-shot forecasts against ARIMA and VAR baselines"
  ),
  Connects_To = c(
    "Cointegration (Section 3), VAR (Section 6)",
    "Volatility clustering (Section 3), GARCH (Section 6)",
    "Week 5 coverage of foundation models and TimeGPT"
  )
)

kable(fw_tbl,
      col.names = c("Extension", "Rationale", "Connects To"),
      caption   = "Table 10: Future Work")
```

---

*Contributions: Daniel Mendoza (Section 3), Jared Maksoud (Section 5), Nicholas Dhaliwal
(Sections 4, 6, 7).*
