---
title: "Crude Oil Forecasting - Models and Results"
subtitle: "ADSP 31006 - Time Series Analysis and Forecasting | Final Project"
author: "Daniel Mendoza, Jared Maksoud, Nicholas Dhaliwal"
date: "March 9th, 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

**Course:** Time Series Analysis and Forecasting
**Instructor:** Prof. Arnab Bose, Ph.D.

This file is self-contained. It loads data, engineers features, and then runs all
models. It is intended as the modeling-focused companion to the full project report.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width  = 9,
  fig.height = 5,
  fig.retina = 2
)
```

# Setup

```{r libraries}
library(tidyverse)
library(lubridate)
library(readxl)
library(gridExtra)
library(knitr)
library(zoo)
library(forecast)
library(tseries)
library(urca)
library(rugarch)
library(FinTS)
library(vars)

# Re-assert dplyr functions masked by MASS (loaded via rugarch/vars)
select <- dplyr::select
filter <- dplyr::filter
lag    <- dplyr::lag
```

```{r load-and-prep, message=FALSE, warning=FALSE}
# --- Data loading ---
x <- read_excel("PET_PRI_SPT_S1_D.xls", sheet = "Data 1", col_names = FALSE)

i   <- which(x[[1]] == "Date")[1]
dat <- x[(i + 1):nrow(x), 1:3]
names(dat) <- c("Date", "USCrudePrice", "EuroCrudePrice")

dnum     <- suppressWarnings(as.numeric(dat$Date))
dat$Date <- ifelse(
  !is.na(dnum) & dnum > 20000,
  as.character(as.Date(dnum, origin = "1899-12-30")),
  as.character(dat$Date)
)
dat$Date <- as.Date(dat$Date, format = "%Y-%m-%d")
bad      <- is.na(dat$Date)
dat$Date[bad] <- as.Date(as.character(dat$Date[bad]), format = "%b %d, %Y")

oil <- dat |>
  mutate(
    USCrudePrice   = suppressWarnings(as.numeric(USCrudePrice)),
    EuroCrudePrice = suppressWarnings(as.numeric(EuroCrudePrice))
  ) |>
  filter(!is.na(Date), !is.na(USCrudePrice), !is.na(EuroCrudePrice)) |>
  arrange(Date)

# --- Feature engineering ---
oil <- oil |>
  arrange(Date) |>
  mutate(
    logWTI          = log(USCrudePrice),
    logBrent        = log(EuroCrudePrice),
    WTI_Return      = c(NA, diff(log(USCrudePrice))),
    Brent_Return    = c(NA, diff(log(EuroCrudePrice))),
    spread          = USCrudePrice - EuroCrudePrice,
    rolling_vol_10  = rollapply(
      c(NA, diff(log(USCrudePrice))),
      width = 10, FUN = sd, fill = NA, align = "right"
    ),
    covid_pulse     = as.integer(
      Date >= as.Date("2020-03-13") & Date <= as.Date("2020-07-01")
    ),
    ukraine_step    = as.integer(Date >= as.Date("2022-02-24")),
    middleeast_step = as.integer(Date >= as.Date("2023-10-06")),
    WTI_sq_return   = c(NA, diff(log(USCrudePrice)))^2
  )

# --- Train/test split ---
h_test <- 63

oil_ret   <- oil |> filter(!is.na(WTI_Return), !is.na(Brent_Return))
n_ret     <- nrow(oil_ret)
train_ret <- oil_ret[1:(n_ret - h_test), ]
test_ret  <- oil_ret[(n_ret - h_test + 1):n_ret, ]

n_lvl     <- nrow(oil)
train_lvl <- oil[1:(n_lvl - h_test), ]
test_lvl  <- oil[(n_lvl - h_test + 1):n_lvl, ]

cat("Train range:", as.character(range(train_ret$Date)), "\n")
cat("Test range :", as.character(range(test_ret$Date)),  "\n")
cat("Test obs   :", h_test, "\n")
```

---

# Intervention Analysis

*Section authored by Jared Maksoud.*

We fit ARIMA with three geopolitical shock dummies as external regressors. Pulse vs. step
classification was determined by inspecting rolling volatility around each event date.
COVID volatility resolved by ~July 2020 (pulse). Ukraine and Middle East volatility stayed
elevated with no reversion (steps).

## Rolling Volatility Around Each Event

```{r int-rolling-vol, fig.height=9}
oil2 <- oil |> filter(!is.na(USCrudePrice), USCrudePrice > 0)

covid_win <- oil2 |>
  filter(Date >= as.Date("2020-01-01"), Date <= as.Date("2020-12-31")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

ukraine_win <- oil2 |>
  filter(Date >= as.Date("2021-12-01"), Date <= as.Date("2022-12-01")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

me_win <- oil2 |>
  filter(Date >= as.Date("2023-06-01"), Date <= as.Date("2024-06-01")) |>
  mutate(vol = rollapply(USCrudePrice, 10, sd, fill = NA, align = "right"))

par(mfrow = c(3, 1), mar = c(3, 4, 3, 1))

plot(covid_win$Date, covid_win$vol, type = "l", col = "black",
     xlab = "", ylab = "10-day rolling sd",
     main = "COVID (2020): volatility resolved ~Jul 2020, treated as pulse")
abline(v = as.Date("2020-03-11"), col = "blue", lty = 2, lwd = 1.5)
abline(v = as.Date("2020-07-01"), col = "blue", lty = 3, lwd = 1.5)

plot(ukraine_win$Date, ukraine_win$vol, type = "l", col = "black",
     xlab = "", ylab = "10-day rolling sd",
     main = "Ukraine (2022): volatility stayed elevated, treated as step")
abline(v = as.Date("2022-02-24"), col = "red", lty = 2, lwd = 1.5)

plot(me_win$Date, me_win$vol, type = "l", col = "black",
     xlab = "Date", ylab = "10-day rolling sd",
     main = "Middle East (2023-24): volatility did not revert, treated as step")
abline(v = as.Date("2023-10-06"), col = "purple", lty = 2, lwd = 1.5)

par(mfrow = c(1, 1))
```

## Intervention ARIMA

```{r int-arima-fit}
y_int <- ts(oil2$USCrudePrice, frequency = 252)

covid_d   <- as.integer(oil2$Date >= as.Date("2020-03-13") &
                        oil2$Date <= as.Date("2020-07-01"))
ukraine_d <- as.integer(oil2$Date >= as.Date("2022-02-24"))
me_d      <- as.integer(oil2$Date >= as.Date("2023-10-06"))

X_int <- cbind(covid        = covid_d,
               ukraine_step = ukraine_d,
               middleeast   = me_d)

fit_int <- auto.arima(
  y_int,
  xreg          = X_int,
  seasonal      = FALSE,
  stepwise      = FALSE,
  approximation = FALSE
)

summary(fit_int)
checkresiduals(fit_int)
```

```{r int-coefs}
beta_int <- coef(fit_int)[c("covid", "ukraine_step", "middleeast")]

cat("\nIntervention Coefficient Estimates:\n")
cat("  COVID pulse effect      :", round(beta_int["covid"],        2),
    "USD/bbl (Mar 13 - Jul 1, 2020)\n")
cat("  Ukraine step effect     :", round(beta_int["ukraine_step"], 2),
    "USD/bbl (permanent from Feb 24, 2022)\n")
cat("  Middle East step effect :", round(beta_int["middleeast"],   2),
    "USD/bbl (permanent from Oct 6, 2023)\n")

int_covid_coef <- round(beta_int["covid"],        2)
int_ukr_coef   <- round(beta_int["ukraine_step"], 2)
int_me_coef    <- round(beta_int["middleeast"],   2)
```

## Counterfactual: Actual vs. Fitted vs. No Interventions

```{r int-counterfactual}
yhat_int   <- as.numeric(fitted(fit_int))
shock_part <- as.numeric(X_int %*% beta_int)
yhat_cf    <- yhat_int - shock_part

plot(oil2$Date, as.numeric(y_int),
     type = "l", col = "black", lwd = 0.7,
     xlab = "Date", ylab = "WTI (USD/bbl)",
     main = "WTI: Actual vs. Fitted vs. No-Intervention Counterfactual")
lines(oil2$Date, yhat_int, col = "blue",      lwd = 1.8)
lines(oil2$Date, yhat_cf,  col = "darkgreen", lwd = 1.8, lty = 2)
abline(v = as.Date("2020-03-13"), col = "gray50", lty = 3)
abline(v = as.Date("2022-02-24"), col = "gray50", lty = 3)
abline(v = as.Date("2023-10-06"), col = "gray50", lty = 3)
legend("topleft",
       legend = c("Actual", "Fitted (with interventions)",
                  "Counterfactual (no interventions)"),
       col = c("black", "blue", "darkgreen"),
       lty = c(1, 1, 2), bty = "n", cex = 0.85)
```

```{r int-contribution}
plot(oil2$Date, shock_part,
     type = "l", col = "darkred", lwd = 1,
     xlab = "Date", ylab = "Intervention contribution (USD/bbl)",
     main = "Total Estimated Intervention Effect Over Time")
abline(h = 0, lty = 2, col = "gray60")
abline(v = as.Date("2020-03-13"), col = "blue",   lty = 3)
abline(v = as.Date("2022-02-24"), col = "red",    lty = 3)
abline(v = as.Date("2023-10-06"), col = "purple", lty = 3)
```

---

# ARIMA on WTI Log Returns

*Section authored by Nicholas Dhaliwal.*

Returns are I(0), confirmed by ADF and KPSS in the full report. We fit ARIMA(p, 0, q)
using the Box-Jenkins procedure: inspect ACF/PACF, run an exhaustive AICc search, then
check neighboring models.

## ACF and PACF

```{r arima-acf}
par(mfrow = c(1, 2))
acf(train_ret$WTI_Return,  lag.max = 40, main = "ACF - WTI Returns (train)")
pacf(train_ret$WTI_Return, lag.max = 40, main = "PACF - WTI Returns (train)")
par(mfrow = c(1, 1))
```

## Model Selection

```{r arima-fit}
wti_ret_ts <- ts(train_ret$WTI_Return, frequency = 252)

arima_fit <- auto.arima(
  wti_ret_ts,
  d             = 0,        # returns are I(0), do not difference
  max.p         = 5,
  max.q         = 5,
  stepwise      = FALSE,    # exhaustive search
  approximation = FALSE,    # exact likelihood
  ic            = "aicc"
)

summary(arima_fit)

p_ar <- arima_fit$arma[1]
q_ma <- arima_fit$arma[2]
cat("\nauto.arima selected: ARIMA(", p_ar, ",0,", q_ma, ")\n", sep = "")
cat("AICc              :", round(arima_fit$aicc,   3), "\n")
cat("Log-likelihood    :", round(arima_fit$loglik, 3), "\n")
```

## Neighboring Model Comparison

```{r arima-neighbors}
cands <- list(
  selected = arima_fit,
  p_up     = Arima(wti_ret_ts, order = c(p_ar + 1, 0, q_ma)),
  q_up     = Arima(wti_ret_ts, order = c(p_ar,     0, q_ma + 1))
)
if (p_ar > 0) cands$p_down <- Arima(wti_ret_ts, order = c(p_ar - 1, 0, q_ma))
if (q_ma > 0) cands$q_down <- Arima(wti_ret_ts, order = c(p_ar,     0, q_ma - 1))

nbr_tbl <- data.frame(
  Model = names(cands),
  AICc  = round(sapply(cands, function(m) m$aicc), 3)
) |> arrange(AICc)

kable(nbr_tbl, caption = "Table 1: ARIMA Neighboring Model AICc Comparison")
```

## Diagnostics

```{r arima-diagnostics}
checkresiduals(arima_fit)
```

```{r arima-ljungbox}
lb_lags <- c(5, 10, 20)
lb_tbl  <- data.frame(Lag = lb_lags, Q_stat = NA_real_, p_value = NA_real_)
for (k in seq_along(lb_lags)) {
  lb <- Box.test(residuals(arima_fit), lag = lb_lags[k],
                 type = "Ljung-Box", fitdf = p_ar + q_ma)
  lb_tbl$Q_stat[k]  <- round(lb$statistic, 3)
  lb_tbl$p_value[k] <- round(lb$p.value,   4)
}
kable(lb_tbl, caption = "Table 2: Ljung-Box Test on ARIMA Residuals")
```

## Forecast and Accuracy

```{r arima-forecast}
arima_fcast <- forecast(arima_fit, h = h_test)

autoplot(arima_fcast) +
  labs(title = paste0("ARIMA(", p_ar, ",0,", q_ma, ") - WTI Return Forecast (",
                      h_test, " days)"),
       x = "Time (training days)", y = "Log Return") +
  theme_minimal()
```

```{r arima-accuracy}
arima_acc <- accuracy(arima_fcast, test_ret$WTI_Return)
kable(round(arima_acc, 6), caption = "Table 3: ARIMA Accuracy - Train and Test")

arima_rmse <- arima_acc["Test set", "RMSE"]
arima_mae  <- arima_acc["Test set", "MAE"]
arima_mase <- arima_acc["Test set", "MASE"]
```

---

# ETS on WTI Price Levels

ETS models price levels directly. The `ets()` function picks the best error/trend/season
specification by AICc. We expect a trend component and no seasonality.

## Model Fit

```{r ets-fit}
wti_lvl_ts <- ts(train_lvl$USCrudePrice, frequency = 252)

ets_fit <- ets(wti_lvl_ts, restrict = FALSE)

summary(ets_fit)
cat("\nSelected:", ets_fit$method, "\n")
cat("AICc    :", round(ets_fit$aicc, 2), "\n")
```

## Diagnostics

```{r ets-diagnostics}
checkresiduals(ets_fit)
```

## Forecast and Accuracy

```{r ets-forecast}
ets_fcast <- forecast(ets_fit, h = h_test)

autoplot(ets_fcast) +
  labs(title = paste0("ETS(", ets_fit$method, ") - WTI Price Level Forecast (",
                      h_test, " days)"),
       x = "Time (training days)", y = "WTI Price (USD/bbl)") +
  theme_minimal()
```

```{r ets-accuracy}
ets_acc <- accuracy(ets_fcast, test_lvl$USCrudePrice)
kable(round(ets_acc, 4), caption = "Table 4: ETS Accuracy - Train and Test")

ets_rmse <- ets_acc["Test set", "RMSE"]
ets_mae  <- ets_acc["Test set", "MAE"]
ets_mase <- ets_acc["Test set", "MASE"]
```

---

# ARMA-GARCH Volatility Model

EDA shows volatility clustering in absolute returns and significant autocorrelation in
squared returns. Both indicate ARCH effects. We confirm this formally with the McLeod-Li
and ARCH-LM tests, then fit GARCH(1,1) with Student-t errors.

## ARCH Pre-Tests

```{r garch-arch-tests}
arima_resid <- residuals(arima_fit)

# McLeod-Li: Ljung-Box on squared residuals
ml_lags <- c(5, 10, 20)
ml_tbl  <- data.frame(Lag = ml_lags, Q_squared = NA_real_, p_value = NA_real_)
for (k in seq_along(ml_lags)) {
  lb_sq <- Box.test(arima_resid^2, lag = ml_lags[k], type = "Ljung-Box")
  ml_tbl$Q_squared[k] <- round(lb_sq$statistic, 3)
  ml_tbl$p_value[k]   <- round(lb_sq$p.value,   6)
}
cat("McLeod-Li Test (Ljung-Box on squared residuals):\n")
kable(ml_tbl, caption = "Table 5: McLeod-Li Test Results")

arch_lm <- ArchTest(arima_resid, lags = 10)
cat("\nARCH-LM Test:\n")
print(arch_lm)
```

## Model Fit

```{r garch-fit}
garch_spec <- ugarchspec(
  variance.model     = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model         = list(armaOrder = c(p_ar, q_ma), include.mean = TRUE),
  distribution.model = "std"
)

garch_fit <- ugarchfit(
  spec   = garch_spec,
  data   = train_ret$WTI_Return,
  solver = "hybrid"
)

show(garch_fit)
```

```{r garch-persistence}
alpha1      <- coef(garch_fit)["alpha1"]
beta1       <- coef(garch_fit)["beta1"]
shape_nu    <- coef(garch_fit)["shape"]
persistence <- alpha1 + beta1
half_life   <- log(0.5) / log(persistence)

cat("\nGARCH(1,1) Key Parameters:\n")
cat("  alpha1 (shock impact) :", round(alpha1,      5), "\n")
cat("  beta1  (persistence)  :", round(beta1,       5), "\n")
cat("  alpha1 + beta1        :", round(persistence, 5), "\n")
cat("  half-life of vol shock:", round(half_life,   1), "trading days\n")
cat("  Student-t dof (nu)    :", round(shape_nu,    2), "\n")

garch_persistence <- round(persistence, 4)
garch_halflife    <- round(half_life,   1)
garch_alpha1      <- round(alpha1, 5)
garch_beta1       <- round(beta1,  5)
```

## Diagnostics

```{r garch-diagnostics, fig.width=11, fig.height=6}
std_resid <- residuals(garch_fit, standardize = TRUE)

par(mfrow = c(2, 2))
acf(std_resid,    lag.max = 30, main = "ACF - Standardized Residuals")
acf(std_resid^2,  lag.max = 30, main = "ACF - Squared Std. Residuals")
pacf(std_resid,   lag.max = 30, main = "PACF - Standardized Residuals")
pacf(std_resid^2, lag.max = 30, main = "PACF - Squared Std. Residuals")
par(mfrow = c(1, 1))
```

```{r garch-lb}
lb_mean_p <- Box.test(std_resid,   lag = 10, type = "Ljung-Box")$p.value
lb_var_p  <- Box.test(std_resid^2, lag = 10, type = "Ljung-Box")$p.value

cat("LB on std. residuals (mean adequacy)  : p =", round(lb_mean_p, 4), "\n")
cat("LB on sq. std. residuals (var adequacy): p =", round(lb_var_p,  4), "\n")
```

## Conditional Volatility

```{r garch-cond-vol}
cond_vol_hist <- sigma(garch_fit)

plot(train_ret$Date, cond_vol_hist,
     type = "l", col = "darkred", lwd = 0.7,
     xlab = "Date", ylab = "Conditional volatility (sigma_t)",
     main = "GARCH(1,1) Estimated Conditional Volatility - Training Set")
abline(v = as.Date("2008-09-15"), col = "orange", lty = 2)
abline(v = as.Date("2020-03-13"), col = "blue",   lty = 2)
abline(v = as.Date("2022-02-24"), col = "red",    lty = 2)
abline(v = as.Date("2023-10-06"), col = "purple", lty = 2)
legend("topright",
       legend = c("sigma_t", "Lehman (2008)", "COVID (2020)",
                  "Ukraine (2022)", "ME (2023)"),
       col = c("darkred", "orange", "blue", "red", "purple"),
       lty = c(1, 2, 2, 2, 2), bty = "n", cex = 0.8)
```

## Volatility Forecast

```{r garch-forecast}
garch_fcast  <- ugarchforecast(garch_fit, n.ahead = h_test)
fcast_vol    <- as.numeric(sigma(garch_fcast))
realized_var <- test_ret$WTI_sq_return

fcast_dates_g <- seq(max(train_ret$Date) + 1, by = "day", length.out = h_test)

plot(fcast_dates_g, fcast_vol,
     type = "l", col = "darkred", lwd = 2,
     xlab = "Date", ylab = "Forecast sigma_t",
     main = paste0("GARCH(1,1) ", h_test, "-Day Volatility Forecast"))

vol_rmse <- sqrt(mean((fcast_vol^2 - realized_var)^2, na.rm = TRUE))
vol_mae  <-      mean( abs(fcast_vol^2 - realized_var),  na.rm = TRUE)

cat("GARCH forecast accuracy (forecast sigma^2 vs. realized r^2):\n")
cat("  RMSE:", round(vol_rmse, 8), "\n")
cat("  MAE :", round(vol_mae,  8), "\n")
```

---

# VAR on WTI and Brent Returns

WTI and Brent have correlation ~0.99 and are cointegrated (confirmed in the full report).
VAR(p) models both return series jointly, using Brent's lags to help forecast WTI.

## Lag Order Selection

```{r var-lag}
ret_mat <- cbind(WTI   = train_ret$WTI_Return,
                 Brent = train_ret$Brent_Return)

var_sel <- VARselect(ret_mat, lag.max = 15, type = "const")
cat("Lag order by criterion:\n")
print(var_sel$selection)

p_var <- var_sel$selection["AIC(n)"]
cat("\nUsing AIC-selected lag p =", p_var, "\n")
```

## Model Estimation

```{r var-fit}
var_fit <- VAR(ret_mat, p = p_var, type = "const")
summary(var_fit)
```

## Granger Causality

```{r var-granger}
gc_b2w <- causality(var_fit, cause = "Brent")
gc_w2b <- causality(var_fit, cause = "WTI")

cat("Granger Causality: Brent -> WTI\n")
print(gc_b2w$Granger)
cat("\nGranger Causality: WTI -> Brent\n")
print(gc_w2b$Granger)

gc_b2w_p <- round(gc_b2w$Granger$p.value, 4)
gc_w2b_p <- round(gc_w2b$Granger$p.value, 4)

cat("\nBrent -> WTI  : p =", gc_b2w_p,
    ifelse(gc_b2w_p < 0.05, "-> Granger-causes WTI", "-> no G-causality"), "\n")
cat("WTI -> Brent  : p =", gc_w2b_p,
    ifelse(gc_w2b_p < 0.05, "-> Granger-causes Brent", "-> no G-causality"), "\n")
```

## Impulse Response Functions

```{r var-irf, fig.width=11, fig.height=5}
irf_b2w <- irf(var_fit, impulse = "Brent", response = "WTI",
               n.ahead = 15, boot = TRUE, ci = 0.95)
irf_w2b <- irf(var_fit, impulse = "WTI",   response = "Brent",
               n.ahead = 15, boot = TRUE, ci = 0.95)

par(mfrow = c(1, 2))
plot(irf_b2w, main = "Brent shock -> WTI response (95% CI)")
plot(irf_w2b, main = "WTI shock -> Brent response (95% CI)")
par(mfrow = c(1, 1))
```

## Forecast and Accuracy

```{r var-forecast}
var_fcast  <- predict(var_fit, n.ahead = h_test)
wti_var_pt <- var_fcast$fcst$WTI[, "fcst"]
wti_var_lo <- var_fcast$fcst$WTI[, "lower"]
wti_var_hi <- var_fcast$fcst$WTI[, "upper"]

fcast_dates_v <- seq(max(train_ret$Date) + 1, by = "day", length.out = h_test)

plot(fcast_dates_v, wti_var_pt,
     type = "l", col = "steelblue", lwd = 2,
     ylim = range(c(wti_var_lo, wti_var_hi)),
     xlab = "Date", ylab = "WTI Log Return",
     main = paste0("VAR(", p_var, ") - WTI Return Forecast (", h_test, " days)"))
polygon(c(fcast_dates_v, rev(fcast_dates_v)),
        c(wti_var_hi, rev(wti_var_lo)),
        col = adjustcolor("steelblue", 0.15), border = NA)
abline(h = 0, lty = 2, col = "gray60")
```

```{r var-accuracy}
var_acc <- accuracy(wti_var_pt, test_ret$WTI_Return)
kable(round(var_acc, 6), caption = "Table 6: VAR Accuracy - WTI Returns (Test Set)")

var_rmse <- var_acc[1, "RMSE"]
var_mae  <- var_acc[1, "MAE"]
var_mase <- var_acc[1, "MASE"]
```

---

# Model Comparison

```{r comparison}
comp_tbl <- data.frame(
  Model  = c("ARIMA", "ETS", "ARMA-GARCH", "VAR", "Intervention ARIMA*"),
  Target = c("WTI log returns", "WTI price (USD/bbl)",
             "Conditional volatility", "WTI + Brent returns",
             "WTI price with shock dummies"),
  RMSE   = c(round(arima_rmse, 6), round(ets_rmse, 4),
             round(vol_rmse,   8), round(var_rmse, 6), NA),
  MAE    = c(round(arima_mae,  6), round(ets_mae,  4),
             round(vol_mae,    8), round(var_mae,  6), NA),
  MASE   = c(round(arima_mase, 4), round(ets_mase, 4),
             NA,                   round(var_mase, 4), NA)
)

kable(comp_tbl,
      col.names = c("Model", "Target", "RMSE", "MAE", "MASE"),
      caption   = "Table 7: Model Comparison",
      na        = "--")
```

*Intervention ARIMA is fitted on the full series for structural estimation. Out-of-sample
accuracy is not applicable.*

ARIMA and VAR both target WTI log returns and can be compared directly on RMSE and MAE.
ETS targets price levels (USD/bbl) so its errors are on a different scale. GARCH targets
conditional variance and is evaluated against realized squared returns.

By use case: ARMA-GARCH is the primary model for volatility and risk management. For
return forecasting, compare VAR vs. ARIMA directly, with VAR preferred when Granger
causality holds. For price level forecasting, use ETS. For shock attribution, read the
intervention ARIMA coefficients directly.

---

*Contributions: Daniel Mendoza (EDA), Jared Maksoud (Intervention Analysis, Section 1),
Nicholas Dhaliwal (Sections 2-5, Model Comparison).*
